---
title: "Remaining Analysis steps for the final project"
output: html_document
author:
  - Piyush Amitabh
---

<!-- Lesson Overview -->

# Topics explored in this pdf_document

1.  **Sexual Dimorphism and Foraging Behavior:** 
We will explore differences in foraging behavior between male and female bees and investigate whether they forage on different plant species. 

2. **Identification of Keystone Plant Species:** Using simulation models, we will assess the potential consequences of the extinction of different plant species, identifying the keystone plant species that are most critical for supporting bee populations. This analysis will involve simulating the removal of these plant species from the ecosystem and evaluating the impact on bee species richness, abundance, and diversity.

3. **Climate and Foraging Patterns:** By integrating climate data, we aim to understand how climatic factors influence bee foraging behavior and whether there are changes in foraging patterns in response to climate variability. This analysis will involve data from sources such as the National Climatic Data Center (NCDC) (https://www.ncdc.noaa.gov/) for climate data (Parmesan et al., 2003).

-------------------------------
```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(igraph)
library(networkD3) # nolint: object_name_linter
library(glue)
```

# Import OBA datasets
```{r}
getwd()
setwd("/home/piyush/Dropbox/Academics/BI510L Data Sci in R/BI510L Data Sci in R_code/bi510l_project")
oba_2018 <- read.csv("data/2018_occurrence.csv")
oba_2019 <- read.csv("data/2019_occurrence.csv")
cat("The colnames are identical:", identical(colnames(oba_2018), colnames(oba_2019)))
oba_2018_2019 <- rbind(oba_2018, oba_2019)
head(oba_2018_2019, 3)
```

# Clean dataset using regex
I will clean and update the associatedTaxa column which is supposed to have the plant name

## 1. Only keep the plant names in the string 
```{r}
sample(oba_2018_2019$associatedTaxa, size = 10)

# using regex to select for the pattern to keep rather than the pattern to drop
# noticed that the plant name is always bounded by the second double quotes ""

# Define the regex pattern
pattern <- '(.*)"(?<plant>.*)"$'

# Use str_match() to extract the match
match <- str_match(oba_2018_2019$associatedTaxa, pattern)
plant_name_list <- match[, "plant"]
plant_name_list[is.na(plant_name_list)] <- ""
oba_2018_2019$associatedTaxa <- plant_name_list
sample(oba_2018_2019$associatedTaxa, size = 10)
```

## 2. Remove all weed names and keep only one plant name
```{r}
cat("We have these many weeds:", sum(grepl("Weedy", oba_2018_2019$associatedTaxa)), "\n")
oba_2018_2019$associatedTaxa <- gsub("Weedy yellow comp.", "", oba_2018_2019$associatedTaxa)
oba_2018_2019$associatedTaxa <- gsub("Weedy", "", oba_2018_2019$associatedTaxa)
cat("Now we have these many weeds:", sum(grepl("Weedy", oba_2018_2019$associatedTaxa)), "\n")

# get rid of ;
cat("We have these many ';'", sum(grepl(";", oba_2018_2019$associatedTaxa)), "\n")
oba_2018_2019$associatedTaxa[grepl(";", oba_2018_2019$associatedTaxa)] <- sapply(
    strsplit(
        oba_2018_2019$associatedTaxa[grepl(";", oba_2018_2019$associatedTaxa)], ";"
    ),
    function(x) x[1]
)
cat("We have these many ';'", sum(grepl(";", oba_2018_2019$associatedTaxa)), "\n")
```

# Add more information

## 1. Add plant resolution
We want to make a new column called plantResolution and assign it to "Family", "Genus" or "Species" depending on the level of resolution associated taxa is resolved to.

```{r}
oba_2018_2019$plantTaxaWordCount <- str_count(oba_2018_2019$associatedTaxa, "\\w+")
head(oba_2018_2019$associatedTaxa, 10)
head(oba_2018_2019$plantTaxaWordCount, 10)
summary(oba_2018_2019$plantTaxaWordCount)
```

```{r}
plant_resolution_list <- rep(NA, nrow(oba_2018_2019))

for (i in seq_along(plant_resolution_list)) {
    word_count <- oba_2018_2019$plantTaxaWordCount[i]
    plant_name <- oba_2018_2019$associatedTaxa[i]

    if (word_count == 1) { # can be genus or family
        if (grepl("aceae$", plant_name)) { # family
            plant_resolution_list[i] <- "family"
        } else {
            plant_resolution_list[i] <- "genus"
        }
    } else if (word_count > 1) { # it is species/subspecies
        plant_resolution_list[i] <- "species"
    }
}
oba_2018_2019$plantResolution <- plant_resolution_list
head(oba_2018_2019$associatedTaxa, 10)
head(oba_2018_2019$plantTaxaWordCount, 10)
head(oba_2018_2019$plantResolution, 10)
```
## 2. Add plant genus
Create a new column called plantGenus that is the genus if the associated taxa was resolved to species or genus, and NA if it was resolved to family. 
```{r}
plant_genus_list <- rep(NA, nrow(oba_2018_2019))

for (i in seq_along(plant_genus_list)) {
    plant_res <- oba_2018_2019$plantResolution[i]
    plant_name <- oba_2018_2019$associatedTaxa[i]

    if (!is.na(plant_res) && (plant_res == "genus")) {
        plant_genus_list[i] <- plant_name
    } else if (!is.na(plant_res) && (plant_res == "species")) {
        plant_genus_list[i] <- strsplit(plant_name, " ")[[1]][1]
    }
}

oba_2018_2019$plantGenus <- plant_genus_list

head(oba_2018_2019$associatedTaxa, 10)
head(oba_2018_2019$plantTaxaWordCount, 10)
head(oba_2018_2019$plantResolution, 10)
head(oba_2018_2019$plantGenus, 10)
```


# Save/Read the new dataframe as csv
## Drop the columns that we are not going to use
```{r}
col_to_save <- c(
    "gbifID", "genericName", "specificEpithet", "species", "sex",
    "taxonRank", "eventDate", "year", "month", "samplingProtocol",
    "associatedTaxa", "county", "decimalLatitude", "decimalLongitude",
    "plantResolution", "plantGenus"
)
oba_2018_2019 <- oba_2018_2019[, col_to_save]
colnames(oba_2018_2019)
```

## Save 
```{r}
write.csv(oba_2018_2019, "data/cleaned_oba_2018_2019.csv", row.names = FALSE)
```
## Read 
```{r}
oba_2018_2019 <- read.csv("data/cleaned_oba_2018_2019.csv")
```
```{r}
# check if all values of column gbifID are unique
unique_gbifID <- length(unique(oba_2018_2019$gbifID))
total_gbifID <- nrow(oba_2018_2019)

if (unique_gbifID == total_gbifID) {
    print("All values of column 'gbifID' are unique.")
} else {
    print("Not all values of column 'gbifID' are unique.")
}
```
This means we can use gbifID as a unique identifier

# 1. Sexual Dimorphism
```{r}
# check the TOTAL PERCENTAGE of MALE vs FEMALE bees caught
total_male <- sum(oba_2018_2019$sex == "MALE", na.rm = TRUE)
total_female <- sum(oba_2018_2019$sex == "FEMALE", na.rm = TRUE)
total_bees <- nrow(oba_2018_2019)

percentage_male <- (total_male / total_bees) * 100
percentage_female <- (total_female / total_bees) * 100

# Create a data frame for the plot
sex_per_df <- data.frame(
    Sex = c("Female", "Male"),
    Percentage = c(percentage_female, percentage_male)
)

# Create a bar plot with ggplot
p_title <- "Percentage of Male and Female Bees"
p <- ggplot(sex_per_df, aes(x = Sex, y = Percentage, fill = Sex)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("coral1", "lightseagreen")) +
    labs(x = "Sex", y = "Percentage") +
    ggtitle(p_title) +
    theme_minimal(18) +
    theme(
        axis.text = element_text(size = 14),
        plot.title = element_text(hjust = 0.5, size = 16)
    )
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)
```
```{r}
# Calculate the percentage of bees caught using different sampling patterns
sampling_patterns <- unique(oba_2018_2019$samplingProtocol)
percentage_per_pattern <- sapply(sampling_patterns, function(pattern) {
    total_bees <- sum(oba_2018_2019$samplingProtocol == pattern, na.rm = TRUE)
    percentage <- (total_bees / nrow(oba_2018_2019)) * 100
    return(percentage)
})

sampling_per_df <- data.frame(
    patterns = sampling_patterns,
    percentage = percentage_per_pattern
)

# Create a stacked bar plot with percentages
p_title <- "Percentage of Bees by Sampling Protocol"
p <- ggplot(sampling_per_df, aes(x = patterns, y = percentage, fill = patterns)) +
    geom_bar(stat = "identity", position = "stack", show.legend = F) +
    labs(x = "Sex", y = "Percentage", fill = "Sampling Protocol") +
    ggtitle(p_title) +
    theme_minimal(18) +
    theme(
        axis.text = element_text(size = 16),
        plot.title = element_text(hjust = 0.5)
    ) +
    scale_fill_brewer(palette = "Set2")
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)
```
```{r}
p_title <- "Percentage of Male and Female Bees by Sampling Protocol"
p <- oba_2018_2019 %>%
    filter(sex != "") %>%
    group_by(sex, samplingProtocol) %>%
    summarise(n_group = n()) %>%
    ungroup() %>%
    group_by(samplingProtocol) %>%
    mutate(
        n_sampling = sum(n_group),
        percentage = n_group / n_sampling
    ) %>%
    ggplot(aes(x = samplingProtocol, y = percentage, fill = sex)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(x = "Sampling Protocol", y = "Percentage", fill = "Sex") +
    ggtitle(p_title) +
    scale_fill_manual(values = c("coral1", "lightseagreen")) +
    scale_y_continuous(labels = scales::percent_format()) +
    theme_minimal(16) +
    theme(
        axis.text = element_text(size = 13),
        plot.title = element_text(hjust = 0.5)
    )
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)

# Generate the plot
p_title <- "Percentage of Bees by Sampling Protocol_2"
p <- oba_2018_2019 %>%
    filter(sex != "") %>%
    group_by(sex, samplingProtocol) %>%
    summarise(n_group = n()) %>%
    ungroup() %>%
    group_by(sex) %>%
    mutate(
        n_sampling = sum(n_group),
        percentage = n_group / n_sampling
    ) %>%
    ggplot(aes(x = sex, y = percentage, fill = samplingProtocol)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(x = "Sex", y = "Percentage", fill = "Sampling Protocol") +
    ggtitle(p_title) +
    scale_fill_brewer(palette = "Set2") +
    scale_y_continuous(labels = scales::percent_format()) +
    theme_minimal(18) +
    theme(
        axis.text = element_text(size = 16),
        plot.title = element_text(hjust = 0.5)
    )
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)
```
# 2. Climate and Foraging Patterns
## Lane county
```{r}
# get climate data for lane county

# Read csv files
avg_temp <- read.csv("data/lane_county_avg_temp_2018_2019.csv", skip = 3, header = 4)
avg_ppt <- read.csv("data/lane_county_precipitation_2018_2019.csv", skip = 3, header = 4)

# rename columns
colnames(avg_temp)[colnames(avg_temp) == "Date"] <- "date"
colnames(avg_temp)[colnames(avg_temp) == "Value"] <- "temp_f"
colnames(avg_ppt)[colnames(avg_ppt) == "Date"] <- "date"
colnames(avg_ppt)[colnames(avg_ppt) == "Value"] <- "ppt_inch"

# add column 'year' and 'month' from the first four string character from the 'Date' column
avg_temp$year <- substr(avg_temp$date, 1, 4)
avg_temp$month <- substr(avg_temp$date, 5, 6)
avg_temp$year <- as.numeric(avg_temp$year)
avg_temp$month <- as.numeric(avg_temp$month)
avg_temp$temp_c <- (avg_temp$temp_f - 32) * (5 / 9) # add celsius
head(avg_temp)
avg_ppt$year <- substr(avg_ppt$date, 1, 4)
avg_ppt$month <- substr(avg_ppt$date, 5, 6)
avg_ppt$year <- as.numeric(avg_ppt$year)
avg_ppt$month <- as.numeric(avg_ppt$month)
avg_ppt$ppt_cm <- avg_ppt$ppt_inch * 2.54 # add cm ppt
head(avg_ppt)
```

```{r}
lane_oba_2018_2019 <- oba_2018_2019 %>%
    filter(county == "Lane")

# find the count of total number of 'gbifID' by year and month in lane_oba_2018_2019
net_count_per_month <- lane_oba_2018_2019 %>%
    group_by(year, month) %>%
    summarise(count = n_distinct(gbifID))

species_count_per_month <- lane_oba_2018_2019 %>%
    filter(species != "") %>%
    group_by(year, month) %>%
    summarise(count = n_distinct(species))

# get and add ppt_cm information from avg_ppt df by year and month
net_count_per_month <- net_count_per_month %>%
    left_join(avg_ppt, by = c("year", "month")) %>%
    left_join(avg_temp, by = c("year", "month"))

species_count_per_month <- species_count_per_month %>%
    left_join(avg_ppt, by = c("year", "month")) %>%
    left_join(avg_temp, by = c("year", "month"))

head(net_count_per_month)
head(species_count_per_month)
```
Plot wrt temp
```{r}
# Scatter plot
ggplot(net_count_per_month) +
    geom_point(aes(x = temp_c, y = count), size = 5, color = "blue") +
    labs(x = "Temperature (°C)", y = "Count") +
    ggtitle("Net Count vs Mean Temp") +
    theme_minimal(18) +
    theme(plot.title = element_text(hjust = 0.5))

ggplot(species_count_per_month) +
    geom_point(aes(x = temp_c, y = count), size = 5, color = "blue") +
    labs(x = "Temperature (°C)", y = "Count") +
    ggtitle("Species Count vs Mean Temp") +
    theme_minimal(18) +
    theme(plot.title = element_text(hjust = 0.5))
```
```{r}
lin_reg <- lm(species_count_per_month$count ~ species_count_per_month$temp_c)
lin_reg_slope <- summary(lin_reg)$coefficients[2]
lin_reg_pval <- summary(lin_reg)$coefficients[7]
summary(lin_reg)
print(lin_reg_slope)
print(lin_reg_pval)


ggplot(species_count_per_month, aes(x = temp_c, y = count)) +
    geom_point(size = 5, color = "blue") +
    labs(x = "Temperature (°C)", y = "Count") +
    ggtitle("Species Count vs Mean Temp") +
    theme_minimal(18) +
    geom_smooth(method = "lm", se = TRUE) +
    theme(plot.title = element_text(hjust = 0.5))
```
Plot wrt ppt
```{r}
# Scatter plot
ggplot(net_count_per_month) +
    geom_point(aes(x = ppt_cm, y = count), size = 5, color = "blue") +
    labs(x = "Precipitation (cm)", y = "Count") +
    ggtitle("Net Count vs Mean Precipitation") +
    theme_minimal(18)+
    theme(plot.title = element_text(hjust = 0.5))


ggplot(species_count_per_month) +
    geom_point(aes(x = ppt_cm, y = count), size = 5, color = "blue") +
    labs(x = "Precipitation (cm)", y = "Count") +
    ggtitle("Species Count vs Mean Precipitation") +
    theme_minimal(18) +
    theme(plot.title = element_text(hjust = 0.5))
```
```{r}
lin_reg <- lm(net_count_per_month$count ~ species_count_per_month$ppt_cm)
lin_reg_slope <- summary(lin_reg)$coefficients[2]
lin_reg_pval <- summary(lin_reg)$coefficients[7]
summary(lin_reg)
print(lin_reg_slope)
print(lin_reg_pval)


ggplot(net_count_per_month, aes(x = ppt_cm, y = count)) +
    geom_point(size = 5, color = "blue") +
    labs(x = "Precipitation (cm)", y = "Count") +
    ggtitle("Net Count vs Mean Temp") +
    theme_minimal(18) +
    geom_smooth(method = "lm", se = TRUE) +
    theme(plot.title = element_text(hjust = 0.5))
```
## Oregon
```{r}
# get climate data for lane county

# Read csv files
avg_temp <- read.csv("data/oregon_avg_temp_2018_2019.csv", skip = 3, header = 4)
avg_ppt <- read.csv("data/oregon_precipitation_2018_2019.csv", skip = 3, header = 4)

# rename columns
colnames(avg_temp)[colnames(avg_temp) == "Date"] <- "date"
colnames(avg_temp)[colnames(avg_temp) == "Value"] <- "temp_f"
colnames(avg_ppt)[colnames(avg_ppt) == "Date"] <- "date"
colnames(avg_ppt)[colnames(avg_ppt) == "Value"] <- "ppt_inch"

# add column 'year' and 'month' from the first four string character from the 'Date' column
avg_temp$year <- substr(avg_temp$date, 1, 4)
avg_temp$month <- substr(avg_temp$date, 5, 6)
avg_temp$year <- as.numeric(avg_temp$year)
avg_temp$month <- as.numeric(avg_temp$month)
avg_temp$temp_c <- (avg_temp$temp_f - 32) * (5 / 9) # add celsius
head(avg_temp)
avg_ppt$year <- substr(avg_ppt$date, 1, 4)
avg_ppt$month <- substr(avg_ppt$date, 5, 6)
avg_ppt$year <- as.numeric(avg_ppt$year)
avg_ppt$month <- as.numeric(avg_ppt$month)
avg_ppt$ppt_cm <- avg_ppt$ppt_inch * 2.54 # add cm ppt
head(avg_ppt)
```

```{r}

# find the count of total number of 'gbifID' by year and month in oba_2018_2019
net_count_per_month <- oba_2018_2019 %>%
    group_by(year, month) %>%
    summarise(count = n_distinct(gbifID))

species_count_per_month <- oba_2018_2019 %>%
    filter(species != "") %>%
    group_by(year, month) %>%
    summarise(count = n_distinct(species))

# get and add ppt_cm information from avg_ppt df by year and month
net_count_per_month <- net_count_per_month %>%
    left_join(avg_ppt, by = c("year", "month")) %>%
    left_join(avg_temp, by = c("year", "month"))

species_count_per_month <- species_count_per_month %>%
    left_join(avg_ppt, by = c("year", "month")) %>%
    left_join(avg_temp, by = c("year", "month"))

head(net_count_per_month)
head(species_count_per_month)
```
Plot wrt temp
```{r}
# Scatter plot
lin_reg <- lm(net_count_per_month$count ~ net_count_per_month$temp_c)
# lin_reg_slope <- summary(lin_reg)$coefficients[2]
# lin_reg_pval <- summary(lin_reg)$coefficients[7]
summary(lin_reg)
p_title = "Oregon - Net Count vs Mean Monthly Temperature"
p <- ggplot(net_count_per_month, aes(x = temp_c, y = count)) +
    geom_point(size = 5, color = "#ff0000") +
    labs(x = "Temperature (°C)", y = "Total #Bees") +
    ggtitle(p_title) +
    theme_minimal(18) +
    geom_smooth(method = "lm", se = TRUE, color="#ff0000") +
    theme(plot.title = element_text(hjust = 0.5))
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)


lin_reg <- lm(species_count_per_month$count ~ species_count_per_month$temp_c)
summary(lin_reg)
p_title <- "Oregon - Species Count vs Mean Monthly Temperature"
p <- ggplot(species_count_per_month, aes(x = temp_c, y = count)) +
    geom_point(size = 5, color = "#ff0000") +
    labs(x = "Temperature (°C)", y = "#Unique Bee Species") +
    ggtitle(p_title) +
    theme_minimal(18) +
    geom_smooth(method = "lm", se = TRUE, color = "#ff0000") +
    theme(plot.title = element_text(hjust = 0.5))
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)
```
Plot wrt ppt
```{r}
lin_reg <- lm(net_count_per_month$count ~ net_count_per_month$ppt_cm)
# lin_reg_slope <- summary(lin_reg)$coefficients[2]
# lin_reg_pval <- summary(lin_reg)$coefficients[7]
summary(lin_reg)
p_title <- "Oregon - Net Count vs Mean Monthly Precipitation"
p <- ggplot(net_count_per_month, aes(x = ppt_cm, y = count)) +
    geom_point(size = 5, color = "lightseagreen") +
    labs(x = "Precipitation (cm)", y = "Total #Bees") +
    ggtitle(p_title) +
    theme_minimal(18) +
    geom_smooth(method = "lm", se = TRUE, color = "lightseagreen") +
    theme(plot.title = element_text(hjust = 0.5))
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)

lin_reg <- lm(species_count_per_month$count ~ species_count_per_month$ppt_cm)
summary(lin_reg)
p_title <- "Oregon - Species Count vs Mean Monthly Precipitation"
p <- ggplot(species_count_per_month, aes(x = ppt_cm, y = count)) +
    geom_point(size = 5, color = "lightseagreen") +
    labs(x = "Precipitation (cm)", y = "#Unique Bee Species") +
    ggtitle(p_title) +
    theme_minimal(18) +
    geom_smooth(method = "lm", se = TRUE, color = "lightseagreen") +
    theme(plot.title = element_text(hjust = 0.5))
p
ggsave(glue("plots/{p_title}.png"), dpi = 300, plot = p)
```
